---
title: "Establishing validity from a computational modeling perspective"
subtitle: "An example focussing on inhibition."
author: "Gidon T. Frischkorn, Ricardo Rebmann, & Klaus Oberauer"
format: html
editor: visual
bibliography: references.json
embed-resources: TRUE
editor_options: 
  chunk_output_type: console
---

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = FALSE,
  output = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  collapse = TRUE
)
```

```{r}
# use relative paths to load & save data
pacman::p_load(here, SimDesign, tidytable, data.table, ggplot2, ComputationalValidity)
```

## Establishing validity from a computational modeling perspective: An example focusing on inhibition

Following commonly used definitions of validity and propositions of validity researchers, establishing the validity of a newly developed measure or test requires its integration in the nomological network of psychological constructs [@cronbachConstructValidityPsychological1955] and providing sufficient evidence for its interpretation and practical use [@messickValidity1987]

## Conflating theoretical with psychometrics questions


## Validity from a computational modeling perspective

### Computational Models implement how processes cause changes in observed behavior


### Identifying suitable compuational models



## Identifying valid behavioral performance measures for inhibition processes

Here we focus on computational models for traditional inhibition tasks such as the Stroop [@stroopStudiesInterferenceSerial1935], Flanker [@eriksenEffectsNoiseLetters1974], or Simon [@simonEffectCompatibilitySR1981] task. These are particularly relevant as the traditional indicator for inhibition from these tasks - that is the difference of reaction times from correct responses between incongruent and congruent conditions- has been criticized for its bad psychometric properties [@draheimReactionTimeDifferential2019; @vonbastianAdvancingUnderstandingIndividual2020]. The main point of criticism was that reaction times are affected by speed-accuracy trade offs and tend to be unreliable, in particular when considering experimental effect measures.

### Simulation I: Diffusion Model for Conflict Tasks

#### Behavioral Effects

```{r avg-behavioral-effects}
dmc_effects_behavior <- dmc_recovery_behavior %>%
  summarize(mean = mean(value),
            sd = sd(value),
            .by = c(nRep, SampleSize, nTrials, measure, indicator))

plot_conditions_behavior <- ggplot(data = dmc_effects_behavior %>% filter (indicator %in% c("comp", "incomp")),
       aes(x = indicator, y = mean, fill = as.factor(nTrials))) +
  facet_grid(SampleSize ~ measure) +
  geom_boxplot() +
  labs(x = "Condition", y = "Performance", fill = "Trials in each\nCondition") +
  theme_gf()

plot_conflictFX_behavior <- ggplot(data = dmc_effects_behavior %>% filter (indicator %in% c("diff")),
       aes(x = indicator, y = sd, fill = as.factor(nTrials))) +
  facet_grid(SampleSize ~ measure) +
  geom_boxplot() +
  labs(x = "Condition", y = "Average Conflict Effect", fill = "Trials in each\nCondition") +
  theme_gf()

ggsave(filename = here("figures","Plot_DMC_Conditions_Behavior.jpg"),
       plot_conditions_behavior, dpi = 300, width = 7, height = 6)
ggsave(filename = here("figures","Plot_DMC_ConflictFX_Behavior.jpg"),
       plot_conflictFX_behavior, dpi = 300, width = 7, height = 6)
```

```{r}
#| label: fig-FX-DMC
#| fig-cap: "Summary of the observed condition differerences and conflict effects in the data simulated from the DMC"
#| fig-subcap:
#|   - "Avg. Performance by Condition"
#|   - "Avg. Conflict Effect"
#| layout-ncol: 2
plot_conditions_behavior
plot_conflictFX_behavior
```

```{r}
dmc_effects_ezDM <- df_ezDM_dmc %>%
  summarize(mean = mean(value),
            sd = sd(value),
            .by = c(nRep, SampleSize, nTrials, measure, indicator))

plot_conditions_ezDM <- ggplot(data = dmc_effects_ezDM %>% filter (indicator %in% c("comp", "incomp")),
       aes(x = indicator, y = mean, fill = as.factor(nTrials))) +
  facet_grid(SampleSize ~ measure) +
  geom_boxplot() +
  labs(x = "Condition", y = "Performance", fill = "Trials in each\nCondition") +
  clean_plot

plot_conflictFX_ezDM <- ggplot(data = dmc_effects_ezDM %>% filter(indicator %in% c("diff")),
       aes(x = indicator, y = sd, fill = as.factor(nTrials))) +
  facet_grid(SampleSize ~ measure) +
  geom_boxplot() +
  labs(x = "Condition", y = "Performance", fill = "Trials in each\nCondition") +
  clean_plot

ggsave(filename = here("figures","Plot_DMC_Conditions_ezDM.jpg"),
       plot_conditions_ezDM, dpi = 300, width = 7, height = 6)
ggsave(filename = here("figures","Plot_DMC_ConflictFX_ezDM.jpg"),
       plot_conflictFX_ezDM, dpi = 300, width = 7, height = 6)
```

```{r}
#| label: fig-FX-ezDM-DMC
#| fig-cap: "Summary of the observed condition differerences and conflict effects in the data simulated from the DMC"
#| fig-subcap:
#|   - "Avg. Performance by Condition"
#|   - "Avg. Conflict Effect"
#| layout-ncol: 2
plot_conditions_ezDM
plot_conflictFX_ezDM
```

```{r merge-effects}
avg_effects_dmc <- rbind(dmc_effects_behavior,dmc_effects_ezDM)

df_reliability_dmc <- df_reliability_dmc %>%
  left_join(avg_effects_dmc)

df_recovery_dmc <- df_recovery_dmc %>%
  left_join(avg_effects_dmc)
```

#### Reliability

```{r}
ggplot(data = df_reliability_dmc,
       aes(x = as.factor(nTrials), y = reliability, fill = as.factor(SampleSize))) +
  facet_grid(indicator ~ measure) +
  ylim(-1,1) +
  geom_boxplot() +
  labs(x = "Number of Trials in each Condition", y = "Estimated Reliability",
       fill = "Sample Size") +
  clean_plot

ggplot(data = df_reliability_dmc %>% filter(indicator == "diff", SampleSize == 100),
       aes(x = sd, y = reliability, color = as.factor(nTrials))) +
  facet_grid(nTrials ~ measure, scales = "free") +
  ylim(-1,1) +
  geom_point(alpha = 0.1, color = "darkgrey") +
  geom_smooth(method = "lm") +
  labs(x = "Standard Deviation of Conflict Effect", y = "Estimated Reliability",
       color = "Number of Trials") +
  clean_plot
```

#### Recovery of DMC parameters

```{r}
ggplot(data = df_recovery_dmc %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("RT","PC")),
       aes(x = genPar, y = rec, fill = as.factor(nTrials))) +
  facet_grid(measure ~ indicator) +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_boxplot() +
  labs(x = "Generating DMC Parameter", y = "Recovery",
       fill = "Sample Size") +
  coord_cartesian(ylim = c(-1,1)) +
  clean_plot

ggplot(data = df_recovery_dmc %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("RT","PC"),
                indicator %in% c("diff")),
       aes(x = mean, y = rec, color = as.factor(SampleSize))) +
  facet_grid(measure ~ genPar) +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  labs(x = "Mean Conflict Effect in the Sample", y = "Recovery",
       fill = "Sample Size") +
  coord_cartesian(ylim = c(-1,1)) +
  clean_plot

ggplot(data = df_recovery_dmc %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("RT","PC"),
                indicator %in% c("diff")),
       aes(x = sd, y = rec, color = as.factor(SampleSize))) +
  facet_grid(measure ~ genPar, scale = "free") +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  labs(x = "SD of Conflict Effect in the Sample", y = "Recovery",
       fill = "Sample Size") +
  coord_cartesian(ylim = c(-1,1)) +
  clean_plot

ggplot(data = df_recovery_dmc %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("v","a","t0")),
       aes(x = genPar, y = rec, fill = as.factor(SampleSize))) +
  facet_grid(measure ~ indicator) +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_boxplot() +
  coord_cartesian(ylim = c(-1,1)) +
  labs(x = "Generating DMC Parameter", y = "Recovery",
       fill = "Sample Size") +
  clean_plot

ggplot(data = df_recovery_dmc %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("RT","PC"),
                SampleSize == 100),
       aes(x = genPar, y = rec_corrected, fill = as.factor(nTrials))) +
  facet_grid(measure ~ indicator) +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_boxplot() +
  ylim(-1,1) +
  coord_cartesian(ylim = c(-1,1)) +
  labs(x = "Generating DMC Parameter", y = "Recovery (Corrected for Reliability)",
       fill = "Sample Size") +
  clean_plot

ggplot(data = df_recovery_dmc %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("v","a","t0"),
                SampleSize == 100),
       aes(x = genPar, y = rec_corrected, fill = as.factor(nTrials))) +
  facet_grid(measure ~ indicator) +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_boxplot() +
  ylim(-1,1) +
  coord_cartesian(ylim = c(-1,1)) +
  labs(x = "Generating DMC Parameter", y = "Recovery (Corrected for Reliability)",
       fill = "Sample Size") +
  clean_plot
```



### Simulation II: Shrinking Spotlight Model of Attention

## 
