---
title: "Validity is a Theoretical Problem: \\\nA Computational Psychometrics Perspective on How to Measure Cognition"
shorttitle: "Validity: A Theoretical Problem"
author:
  - name: Gidon T. Frischkorn
    orcid: 0000-0002-5055-9764
    email: gidon.frischkorn@psychologie.uzh.ch
    corresponding: true
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
  - name: Ricardo Rebmann
    orcid: 0009-0005-5164-3211
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
      - name: Singular IT GmbH
  - name: Klaus Oberauer
    orcid: 0000-0003-3902-7318
    affiliations:
      - name: University of Zurich
        department: Department of Psychology
author-note:
  disclosures:
    study-registration: "This study was not preregistered."
    data-sharing: "Analysis code and simulation scripts are available at https://github.com/GidonFrischkorn/ComputationalValidity. Simulation data and results are archived on the Open Science Framework at https://doi.org/10.17605/OSF.IO/6QNRV."
    conflict-of-interest: "Ricardo Rebmann is employed by Singular IT GmbH. The employer had no role in the conceptualization, design, analysis, or interpretation of the research, nor in the writing of the manuscript. The remaining authors declare no conflicts of interest."
    financial-support: "This research was supported by a grant from the Swiss National Science Foundation (SNSF) awarded to Gidon T. Frischkorn (Grant No. 208828). Ricardo Rebmann's contribution was partly supported by Singular IT GmbH through allocated work time."
    authorship-agreements:
      "Gidon T. Frischkorn: Conceptualization, Methodology, Software, Formal Analysis, Visualization, Writing -- Original Draft.
      Ricardo Rebmann: Methodology, Software, Formal Analysis, Writing -- Review & Editing.
      Klaus Oberauer: Conceptualization, Supervision, Writing -- Review & Editing."
format:
  apaquarto-docx: default
  apaquarto-pdf:
    keep-tex: true
floatsintext: true
output-dir: "."
bibliography: validity_computational_psychometrics.bib
abstract: "Psychometric methods alone cannot determine what cognitive tasks measure. We argue that validity – whether behavioral indicators reflect the cognitive processes they are intended to measure – is fundamentally a theoretical problem. Traditional validation through convergent and discriminant correlations is circular when constructs are defined through the same correlational patterns used to validate them. Computational cognitive models can break this circularity by specifying how cognitive processes generate behavior, allowing validity to be assessed independently for each indicator through simulation. We demonstrate this framework using two models of conflict processing – the Diffusion Model for Conflict (DMC) and the Shrinking Spotlight model (SSP) – applied to tasks commonly used to measure attention control. Simulations reveal four measurement phenomena invisible to correlational validation: (1) process impurity, where many behavioral indicators conflate multiple cognitive processes; (2) differential validity, where RT difference scores uniquely reflect attention-control parameters under both models; (3) reliability-validity dissociations, where highly reliable mean scores show weaker validity for attention control than less reliable difference scores; and (4) correlation transfer failure, where process-level correlations across tasks transfer to indicator-level correlations only to the extent that indicators validly reflect the correlated parameters. These phenomena emerge consistently from both models despite different processing assumptions. Our work shows that advancing validity requires explicit theories of how cognition generates behavior, not more sophisticated psychometric methods or models."
keywords:
  - validity
  - computational modeling
  - individual differences
  - attention control
  - psychometrics
execute:
  echo: false
  eval: false
  warning: false
  message: false
---

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = FALSE,
  output = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  collapse = TRUE
)
```

```{r}
# use relative paths to load & save data
pacman::p_load(here, SimDesign, tidytable, data.table, ggplot2, ComputationalValidity)
# source(here("scripts","LoadResultsFiles.R"))
```

The present work is concerned with a problem that cuts across cognitive psychology: Psychometric methods alone cannot determine what cognitive tasks measure, and decades of methodological refinement have not resolved it. We argue that the problem of validity – ensuring that the observed variables we obtain from task performance measure what we intend to measure – cannot be solved by refining methods alone. Validity rests on theoretical commitments, and therefore, ensuring validity requires theoretical work [@Borsboom2004]. We develop our argument with reference to several example cases, with a primary focus on the measurement of attention control ability.

Across several domains, researchers have tried to establish behavioral indicators that measure latent attributes of the cognitive system. Here and throughout, we use *behavioral indicators* to refer to summary statistics computed from trial-level behavioral data in cognitive tasks – such as mean reaction time, accuracy rates, or difference scores – that researchers typically use to quantify individual differences.[^1] Researchers have tried to work out what behavioral indicators from cognitive tasks measure by (1) optimizing their reliability and (2) assessing validity through convergent and discriminant correlations among indicators.

[^1]: We will sometimes simply refer to "task" as shorthand for "a behavioral indicator obtained from behavior in a task."

In attention control research, researchers used experimental paradigms for studying cognitive conflict, such as the Stroop task and the flanker task, to measure individual differences in attention-control (AC) ability. Although conflict tasks produce robust experimental effects, the size of these effects, measured through performance differences between experimental conditions, often fails to correlate substantially across individuals, making it difficult to establish a coherent factor of AC ability [@ReyMermetEtAl2018; @EnkaviEtAl2019]. This is in part due to the limited reliability of difference scores – the so-called reliability paradox [@HedgeEtAl2018]. Some researchers responded to this problem by designing new conflict tasks, replacing difference scores by raw performance scores to improve the reliability of AC measures, and thereby succeeded in demonstrating coherent factor structures [@DraheimEtAl2021; @Burgoyne2023]. Others argued that these new measures do not tap the same processes as the traditional ones, that their improved coherence may reflect shared non-AC processes rather than a common AC ability [@ReyMermet2025]. Additionally, there is some evidence that attention control factors derived from traditional tasks are indistinguishable from factors reflecting processing speed [@LoefflerEtAl2024].

A similar pattern characterizes working memory research. Complex span and N-back tasks are both assumed to reflect working-memory capacity (WMC), but they correlate only weakly despite adequate reliability, a pattern first documented by @KaneEtAl2007 and confirmed meta-analytically by @RedickLindsey2013. The resulting debate spans nearly two decades: Some researchers argue that the tasks measure distinct constructs [@KaneEtAl2007; @BurgoyneEtAl2024WM], whereas others maintain that WMC is a unified ability complicated by task-specific demands [@SchmiedekEtAl2009; @WilhelmEtAl2013; @SchmiedekEtAl2014; @WilhelmEtAl2025].

The processing speed literature tells a complementary story. Measures of speed in simple tasks are substantially correlated with each other, supporting a latent factor commonly labeled as "processing speed". This label motivated the question: The speed of what process(es) is reflected in the factor? Early attempts to isolate process-specific speed using elementary cognitive tasks such as memory-scanning [@Sternberg1966] or the Hick paradigm [@Hick1952] failed to produce robust correlations with intelligence [@BeauducelBrocke1993]. @Lohman1994 explained why: Most between-person variance is carried by the intercept of regression models, which reflects what is common across conditions, not by slopes or experimental contrasts. Hence, traditional measures of "processing speed" conflate the speed of an unknown mixture of multiple processes; attempts to isolate specific processes have been largely abandoned because they did not fulfil the promise of illuminating the cognitive foundations of intelligence.

Across these domains, the same conceptual problem recurs. High reliability and well-converging factor models using behavioral indicators for measuring cognitive processes do not guarantee valid measurement of the intended constructs. Factor coherence demonstrates shared variance, but not what mechanisms produce it. Correlation patterns cannot adjudicate whether weak convergent validity reflects measurement inadequacy or genuine construct distinctiveness. Many researchers treat these difficulties as methodological problems requiring better psychometric techniques, larger samples, or more careful task selection. We argue that the core problem is theoretical. Without explicit commitments about how cognitive processes generate observed behavior, validation becomes circular: Behavioral indicators from one task are validated by their correlations with indicators from other tasks assumed to measure the same construct; the validity of which in turn rests on their correlation with other tasks.

We develop this argument by building on @Borsboom2004, who defined validity in explicitly causal terms: "A test is valid for measuring an attribute if and only if (a) the attribute exists and (b) variations in the attribute causally produce variations in the outcomes of the measurement procedure" (p. 1061). We propose to use computational cognitive models to implement this framework: They specify what cognitive processes exist (as model parameters) and how these processes generate observations (through explicit generative mechanisms). To the extent that we established confidence in the adequacy of a cognitive model for a task, we can use it to inform the validity of behavioral indicators from that task as measurements of the processes that the model describes. We can then use simulations with the model to test the causal link, that is, whether individual differences in behavioral indicators reflect differences in the generating processes.

## Classical Validity Theory

Theory has been central to validity frameworks since their inception, but the specific theoretical requirements for cognitive task validation have often remained implicit in practice. @CronbachMeehl1955 introduced the concept of construct validity for situations where the measured attribute is not directly observable. They argued that validation is fundamentally theory testing: The construct is embedded in a "nomological network" that generates testable predictions about patterns among measures, experimental manipulations, and group differences. Critically, they recognized the risk of circularity: "Unless the network makes contact with observations, and exhibits explicit, public steps of inference, construct validation cannot be claimed" (p. 290).

@CampbellFiske1959 operationalized this logic through the multitrait-multimethod (MTMM) matrix, separating trait variance from method variance and establishing convergent and discriminant validity as key criteria.[^2] However, the MTMM framework embeds a critical assumption: That correlation magnitude can adjudicate whether different measures target the same underlying construct. This becomes problematic when the processes reflected in behavioral indicators and generating observed correlations are complex or incompletely understood.

[^2]: The logic underlying the MTMM matrix – that consistent covariation among behavioral indicators reveals common source traits more fundamental than surface correlations – originates in @Cattell1943b's factor-analytic approach to personality, where factor analysis of trait ratings identified source traits underlying clusters of covarying surface behaviors [@Cattell1943a]. @Cattell1944 later formalized this reasoning as the principle of "parallel proportional profiles": if the same factor structure replicates across independent samples or data types (e.g., self-report vs. observational data), this constitutes evidence for a real underlying trait. In practice, comprehensive tests of factorial invariance across observational contexts are rarely conducted, and recent research indicates that within-person covariance structures can differ considerably from between-person structures [@Molenaar2004; @Hamaker2005], violating the assumption of parallel profiles.

Later refinements in the conception of validity maintained the central role of theory. @Messick1995 argued for a unified conception centered on "score meaning," emphasizing that the substantive component of validity – understanding what processes underlie task performance – is essential to interpretation. @Kane2013 developed an argument-based framework in which researchers must construct explicit arguments for the intended interpretation or use of test scores, and each argument's assumptions can be evaluated empirically; in principle, such arguments could incorporate process models as warrants for construct interpretations. Both frameworks thus recognize the need for theoretical grounding of score meaning [see @MarkusBorsboom2013 for a comprehensive treatment]. However, neither specifies what form such theoretical commitments must take for cognitive constructs – what level of mechanistic detail is required, or how process-to-indicator mappings should be formalized. This gap is consequential: without explicit guidance on the form of process specification, validation efforts default to correlational evidence, which, as we argue next, introduces circularity.

@Whitely1983 [subsequently Embretson] drew a distinction between *construct representation*, which refers to the theoretical assumptions about mechanisms that generate task performance, and *nomothetic span*, referring to the correlational network among measures. She argued that psychology had over-emphasized nomothetic span at the expense of construct representation, focusing on whether tasks correlate rather than on what processes produce performance, and that construct representation – understanding what processes generate performance – is just as important as nomothetic span – establishing correlations with other measures. This distinction is especially consequential for cognitive psychology, where process models can specify construct representation with some precision, and it foreshadows our central argument.

Despite consistently emphasizing theoretical commitments – nomological networks, trait-method distinctions, score meaning, construct representation, causal mechanisms – classical frameworks have never specified how to establish validity of behavioral indicators in practice. Our critique, therefore, is not that these frameworks fail to recognize the importance of theory – they do – but that they do not provide the tools to translate theoretical commitments into concrete validity assessments for cognitive indicators. Without explicit process theories, validation efforts rely on convergent validity: The expectation that indicators assumed to measure the same construct should correlate. As we show next, this logic becomes circular when the construct itself is defined through indicator correlations.

## The Circularity Problem in Correlation-Based Validation

The validation logic underlying the research programs described above is often circular: When constructs are defined through task correlations, those correlations cannot also serve as evidence for construct validity. This limitation is structural, not a shortcoming of individual research programs, which typically exemplify current best practices.

Consider @DraheimEtAl2021's effort to improve attention control measurement by adapting conflict tasks with response deadlines while avoiding difference scores [@Draheim2019]. They validated these new tasks against standard psychometric benchmarks: High reliability, strong correlation with other attention control tasks, and substantial correlations with working memory capacity and fluid intelligence. On these criteria, the tasks succeeded, leading the authors to conclude they had achieved improved measurement of attention control.[^3] Yet this validation strategy embeds a circular logic. @ReyMermetEtAl2018 and @ReyMermet2025 have argued that AC ability is probably not a unitary construct. Whether or not the ability to control attention in different situations reflects in part a general AC ability is a substantive question, and to resolve it we need valid measures of attention control. Yet, the approach of Draheim and colleagues for validating their new measures presupposes the existence of a general AC ability. By selecting tasks based on intercorrelations with other tasks *assumed* to measure AC, the approach biases the substantive question it aims to answer [@Oberauer2024]. The same circularity characterizes working memory research, where @WilhelmEtAl2013 and @WilhelmEtAl2025 defend capacity as a unified construct and question the validity of N-back for measuring it, whereas @BurgoyneEtAl2024WM claim that span and N-back tasks measure distinct abilities: The inference that N-back performance has poor validity for measuring working-memory capacity rests on the assumption that working-memory capacity is a unified construct. Conversely, the inference that working-memory capacity is not unified rests on the assumption that N-back performance is a valid measure of working-memory capacity.

[^3]: Similarly, @Yanguez2024's approach to improve executive function measurement by applying computational models to improve convergence and stability of executive function models also relies on the same circular logic: They select indicators based on their correlations with other indicators assumed to measure executive function, and then claim that these correlations validate the indicators. This approach presupposes that the indicators measure a common construct, which is the very question at stake.

No amount of psychometric refinement can resolve the interpretive ambiguity this creates. Does factor coherence among behavioral indicators from conflict tasks demonstrate shared control processes, or merely shared evidence-accumulation efficiency [@LoefflerEtAl2024]? When working memory tasks correlate only moderately with each other, does this reflect distinct constructs or inadequate indicators of a common one? Correlational evidence alone cannot adjudicate, because the validation logic itself is circular.

The circularity becomes apparent when we trace the reasoning explicitly (see the left panel of @fig-circularity for an illustration): (1) We theorize that there is a general AC ability. To test this assumption we need tasks that validly measure AC. (2) From the assumption of a general ability we predict that tasks that require AC to a substantial degree should be positively correlated. (3) We sample several tasks of which we believe that they require AC, for instance Anti-Saccade and Visual Arrays tasks, and find performance in these tasks to be positively correlated. (4) We conclude that these tasks are valid measures of AC, and (5) we conclude that their positive correlation confirms the assumption of a general AC ability. The reasoning is circular because the validation of the tasks rests on the substantive assumption that we set out to test with these tasks. The same circular reasoning applies to working memory (correlations between complex span and N-back performance) and processing speed (RT-intelligence correlations). As @Borsboom2004 argued, this creates a closed circle of test interpretation where tests are validated by their intercorrelations, and the intercorrelations are explained by appealing to the constructs that were posited to explain those intercorrelations in the first place. Factor analysis does not break this circle – extracting a latent factor demonstrates shared variance, but the psychological meaning of that factor depends on assumptions about what the tasks measure [@Borsboom2005; @MarkusBorsboom2013]. An "attention control" factor might equally reflect general processing speed; a "working-memory capacity" factor might reflect executive attention rather than storage capacity. Without independent specification of what processes generate the factor, construct labels merely relabel the correlation pattern.

As @Whitely1983 recognized, nomothetic span (the correlational network among measures) cannot substitute for construct representation (the theoretical account of how processes generate performance). When tasks fail to correlate as expected – when attention-control factors relate more strongly to speed than to working memory [@LoefflerEtAl2024] – correlation patterns alone cannot adjudicate whether the construct is wrong or the measurement is invalid.

![Schematic contrast between correlation-based validity (left, five-step cycle) and the computational approach (right, five-step open chain). *Left*: The reasoning proceeds clockwise through steps 1–4 – a general ability is assumed, used to predict task correlations, task correlations are observed, and those correlations are cited as evidence for indicator validity. The curved arrow from step 3 to step 5 marks the double-duty problem: the same correlational evidence simultaneously validates the indicators (step 4) and confirms the original ability assumption (step 5). The dashed arrow marks this closing link as the problematic step that makes the reasoning circular and non-falsifiable. *Right*: The computational approach shares the same five-step structure but does not close the loop. A formal theory is specified (1), cognitive processes are formalized as model parameters (2), data are generated from the model (3), and indicator validity is assessed via simulation (4). Step 5 – cross-task correlations – becomes an open empirical prediction rather than a definitional assumption, because the validity of each indicator has been established independently. The lightly shaded box ("Reproduce empirical effects") represents an additional validation step in which simulated data from the model are checked against well-established qualitative patterns of empirical data; this model adequacy check constitutes a prerequisite before parameter-level validity can be meaningfully assessed.](../figures/manuscript/Circularity_Diagram.png){#fig-circularity}

Breaking this circularity requires establishing validity independently for each behavioral indicator instead of using cross-task correlations as validity evidence. The validity of an indicator must be demonstrated through independent specification of how variation in the target attribute produces variation in that indicator (see right panel of @fig-circularity), not through correlations with other indicators assumed to measure the same construct. Only after validity has been established independently for indicators from different tasks can their inter-correlations test whether performance reflects a unitary ability or diversity in supposedly related processes. This sequence reverses current practice: Rather than selecting tasks based on their correlations and then claiming those correlations validate the construct, researchers must first specify what processes each task engages and which indicators reflect those processes, then test whether predicted process-level relationships emerge.

Concretely, independent indicator validation requires a *generative* specification: A model that predicts how individual differences in a target process map onto individual differences in an observable indicator [@KvamAlaukik2025], without presupposing the correlational patterns this prediction aims to explain. Verbal theories can identify candidate processes, but they cannot determine how strongly each process contributes to a given indicator, or whether nuisance processes contaminate the indicator. Quantifying the process-to-indicator mapping requires progressively more formal specification – from verbal frameworks that constrain interpretation, through semi-formal decompositions that test component processes, to generative computational models that simulate complete data from explicit process assumptions. This formalization continuum defines what kinds of validity claims each level of theory can support.[^4]

[^4]: To be precise: What we aim to measure is not a process itself, but an attribute of a mechanism or process – represented by a parameter of a model – that varies across individuals. Hence, validity requires showing that variation in that attribute causally produces variation in the indicator. We continue to speak of "processes generating indicators" as shorthand, but the target of measurement is always a quantitative attribute of a process, not the process itself.

## From Ability Constructs to Process Parameters

The independent validation of behavioral indicators required by the preceding argument demands a specific theoretical move: Reconceptualizing cognitive constructs as information-processing mechanisms with specifiable behavioral consequences, rather than as latent factors defined through shared variance. Although this is common in cognitive experimental psychology, in most individual-differences research, cognitive constructs are treated as latent factors – unobserved variables that account for shared variance among mostly behavioral indicators. This framing naturally leads to validation through convergent validity, and thus back into the circularity diagnosed above. 

A process-oriented view [@FrischkornEtAl2022] begins from a different starting point: Cognitive constructs are hypotheses about information-processing mechanisms that generate behavior. From this perspective, "attention control" refers not to an ability – an unspecified source of variance and covariance between individuals – but to a set of mechanisms and processes by which the cognitive system controls attention, and which vary in their effectiveness and efficiency between individuals [@Oberauer2024]. These processes include resolving conflict, enhancing relevant and suppressing irrelevant information, flexibly adjusting processing strategies, with computational and temporal properties that can be specified with varying degrees of formality [e.g., @BotvinickEtAl2001; @CohenEtAl1990; @HubnerEtAl2010; @LoganGordon2001; @Oberauer2024; for a comprehensive overview, see @ReyMermet2025]. Critically, tasks do not "measure" constructs directly – they create conditions under which processes operate and produce observable outcomes. A flanker task does not measure attention control in the way a thermometer measures temperature. It creates a situation where control processes contribute to determining reaction times and error rates in the different experimental conditions. But as we will see from the simulation results presented later, different behavioral indicators capture different aspects of the processes that interplay and generate the observed behavior. In sum, how strongly a specific behavioral indicator indexes a specific process depends on the processing architecture [@TownsendAshby1983]: What processes are engaged, how they interact, and how their activity maps onto behavior.

Under this view and following the conception of validity proposed by @Borsboom2004, a behavioral indicator is valid for a cognitive process to the extent that variation in that process causally influences variation in the indicator. This has several implications. Validity is *graded*: An indicator can be more or less valid depending on how strongly the target process influences it. It is *parameter-specific*: Behavioral indicators from the same task may vary in their validity for specific model parameters. It is *model-relative*: Validity claims depend on the theoretical assumptions embodied in a cognitive model, a point we return to in the Discussion. And it is *testable*: Predictions from the model can be evaluated against data.

The key point is that, following @Borsboom2004, validity claims require explicit theoretical commitments about how processes generate behavior. Without such commitments, validation becomes circular; with them, it becomes testable.

## Validation Through Generative Models

The strongest validity assessments come from formal *generative* models of cognitive processes that specify process parameters and their interplay mathematically, making all assumptions about the processes underlying behavior explicit [@GuestMartin2021]. In such a model, individual differences in constructs correspond to individual differences in model parameters [@Oberauer2016]. This idea underlies what has been termed *cognitive psychometrics* – locating individual differences at the level of process parameters rather than behavioral scores [@FrischkornEtAl2022; @RouderLu2005; @SchubertFrischkorn2020; @Vandekerckhove2011].

Before using a generative model to assess indicator validity, its adequacy must be established: The model should reproduce well-established qualitative patterns in empirical data (e.g., congruency effects, RT distribution shapes; see the shaded box in @fig-circularity). This adequacy check can be complemented by fitting the model to empirical data, which also enables direct parameter estimation as an alternative route to measurement (see the Discussion for details).

A formal generative model specifies all processes and parameters necessary to simulate behavioral data trial by trial. If such a model correctly describes how cognitive processes generate behavior, and if a behavioral indicator reflects a specific process, then individual differences in the indicator should track individual differences in the corresponding model parameter. We can assess this through four steps: 1) specify a generative model formalizing theory about how behavior in a cognitive task arises, 2) simulate data with parameter values that vary across simulated participants, 3) compute the same behavioral indicators researchers typically use (e.g., mean RT, accuracy, RT difference scores) from those data, and then 4) assess how strongly those indicators correlate with the parameter values that researchers aim to measure (see the dotted arrow in the right panel of @fig-circularity). Strong correlations indicate valid measurement of the corresponding process; weak correlations indicate confounding of the behavioral indicator, insensitivity of the indicator to the process of interest, or a non-linear relationship between process variation and indicator variation.

Because the validation criterion – the generating parameter values – is established independently of between-task correlations, validity claims no longer rest on the circular logic diagnosed above. They remain conditional on model adequacy, but model adequacy can be evaluated independently through model fitting and comparison techniques and through experimental tests. At a minimum, the generative model should reproduce well-established qualitative patterns from empirical data. In the case of models for conflict tasks these are, among others, the congruency costs (faster and more accurate responses in congruent than incongruent conditions), the characteristic range and shape of reaction-time distributions, fast-error effects, and the positive or negative slope of delta plots. Both the DMC [@UlrichEtAl2015] and the SSP [@WhiteEtAl2011; @WhiteEtAl2018] that we use to illustrate the proposed validation framework for behavioral indicators gain credibility from the fact that their simulated data reproduce these qualitative regularities of conflict-task performance. A model that fails to reproduce these basic regularities cannot yield trustworthy validity estimates for its parameters. Although the present paper focuses on indicator validity via simulation-based validity assessment, this empirical adequacy check constitutes a prerequisite that should be satisfied before parameter-level validity can be meaningfully assessed.

## An Application to the Measurement of Attention Control

### Simulation Design

We use cognitive conflict tasks (such as flanker, Simon, and Stroop tasks) as an example for illustrating how validity of indicators for measuring individual differences in AC can be established based on formal generative models. From our perspective, these tasks are well suited as a testbed, because their traditional behavioral indicators – RT differences between incongruent and congruent conditions – exhibit the validity challenges discussed above: Several studies found weak and inconsistent correlations of RT differences between tasks [@EnkaviEtAl2019; @FrischkornBastian2021; @ReyMermetEtAl2018]. Some authors have interpreted this finding as evidence against a general AC ability [@ReyMermetEtAl2018; @ReyMermet2025], whereas others have criticized RT differences as invalid measures of AC [@Draheim2019; @Yanguez2024], illustrating the ambiguity that results from the circularity in the convergent validation program. Nevertheless, RT differences remain widely used in individual-differences research [@DraheimEtAl2021; @FriedmanMiyake2004]. Several generative computational models exist for conflict tasks [@UlrichEtAl2015; @WhiteEtAl2011], and they can be used to simulate trial-level data from which multiple candidate indicators can be computed.

We focus on two generative models formalizing different perspectives on conflict processing. The Diffusion Model for Conflict [DMC; @UlrichEtAl2015] conceptualizes conflict as arising between automatic and controlled evidence accumulation. *Automatic evidence accumulation* is triggered by salient but task-irrelevant features, such as the word meaning in a Stroop task. This automatically accumulated evidence surges initially and is subsequently suppressed, resulting in a temporary activation bump of irrelevant evidence that is described by its amplitude *A*, and its temporal dynamics of decline $\tau$. *Controlled processing* accumulates evidence from task-relevant information with a constant drift rate $\mu_c$. At any time during task processing the two sources of evidence combine to drive a diffusion process toward one of two response boundaries (set by the boundary-separation parameter *b*), with non-decision time ($T_{er}$) for the summed time of encoding and motor execution. In the DMC, the efficiency of attention control is represented in the magnitude of automatic activation, quantified by its amplitude *A* and its rate of decline, $\tau$.

The Shrinking Spotlight model [SSP; @WhiteEtAl2011; @WhiteEtAl2018] was developed specifically for the flanker task and describes conflict resolution as a gradual concentration of the attentional focus on the relevant information. Attention initially encompasses both relevant and irrelevant information (e.g., the target and the flankers), then gradually narrows onto task-relevant information (e.g., the identity of the target). Parameters include perceptual input strength (*P*), initial spread of the attentional spotlight ($sd_0$), the rate at which the spotlight narrows ($r_d$), boundary separation (*b*), and non-decision time ($T_{er}$). The efficiency of attention control is represented by how much irrelevant information is processed, which is determined by the initial spread of the attentional focus and its rate of narrowing. Because $sd_0$ and $r_d$ jointly determine how quickly flanker interference diminishes, they exhibit strong tradeoffs that compromise joint identifiability [@WhiteEtAl2018]; we therefore held $r_d$ fixed and varied only $sd_0$ as the attention-control parameter in our simulations. Using two architecturally distinct models allows us to distinguish validity patterns that are architectural-general – emerging from both models despite different processing assumptions – from those that are model-specific. @fig-model-architecture illustrates the core processing mechanism of each model; @tbl-parameters gives a summary of the model parameters and their interpretation.

![Conceptual illustration of the two generative models used in simulations. **A (DMC, top left):** Mean evidence accumulation paths toward the correct response boundary (*a*) for congruent (blue solid) and incongruent (red dashed) conditions, starting at *z* = *a*/2. The dot-dashed line shows the pure controlled-drift asymptote $\mu_c$. The horizontal bracket marks the RT difference score. **A2 (DMC, bottom left):** Automatic activation component $\mu_A$ characterised by the peak amplitude $A$ and the time $tau$ to reach this peak for both conditions (congruent: +, incongruent: −). **B (SSP, right):** Attentional spotlight profiles (Gaussian weight over spatial position) at four successive time points. The solid vertical line marks the target; dashed lines mark flanker positions. The broad initial spread (sd~0~, annotated in *t*~1~) gives flankers substantial weight ("Flankers dominate"); the spotlight narrows at the shrink rate $r_d$ (annotated in *t*~2~ as the reduction from the initial width). The attentional weight for each stimulus in the spotlight modulates the perceptual evidence strength $P$. As the spotlight narrows the target dominates, reversing the net drift direction.](../figures/manuscript/Model_Architecture.png){#fig-model-architecture}

| DMC | Interpretation | SSP | Interpretation |
|-----|---------------|-----|---------------|
| $\mu_c$, drift rate | Efficiency of accumulating evidence from the relevant feature dimension | *P*, perceptual input strength | Efficiency of accumulating information from features within the attentional focus |
| *b*, boundary separation | Amount of evidence required for a decision; degree of caution | *b*, boundary separation | Amount of evidence required for a decision; degree of caution |
| *A*, amplitude | Peak strength of automatic activation triggered by the irrelevant feature | $sd_0$, initial spotlight width | Initial width of the attentional focus; broader spread gives more weight to irrelevant features |
| $\tau$, time to peak | Speed at which automatic activation rises and decays | $r_d$, shrink rate^a^ | Rate at which the attentional spotlight narrows onto the target |
| $T_{er}$, non-decision time | Summed time for encoding and motor execution | $T_{er}$, non-decision time | Summed time for encoding and motor execution |

: Summary of model parameters and their interpretation for the Diffusion Model for Conflict (DMC) and the Shrinking Spotlight model (SSP). ^a^Held fixed in simulations due to strong tradeoffs with $sd_0$ that compromise joint identifiability. {#tbl-parameters}

We conducted separate simulations of behavior in a conflict task using DMC and SSP, following identical design principles. The task was to make binary decisions on a relevant stimulus dimension (e.g., the feature of the target of a flanker task) in the context of an irrelevant stimulus dimension (e.g., the feature of the flankers). Half of the trials were congruent (i.e., the flanker feature matching the target feature) and the other half were incongruent. We systematically varied sample size (*N* = 25, 50, 100, 200) and trials per condition (50, 100, 200), with each design cell replicated 250 times. For each simulated participant, we drew parameter values from distributions producing realistic behavioral patterns (see Appendix), and generated trial-level RT and accuracy data. From these data we computed standard behavioral indicators: mean scores (average RT and accuracy) and difference scores (RT or error rate in incongruent minus congruent trials). Because researchers have argued that applying computational measurement models to estimate parameters improves measurement [@Yanguez2024], we also calculated EZ-Diffusion Model parameters [@WagenmakersEtAl2007] – a simplified cognitive model that estimates basic diffusion parameters from aggregate behavioral statistics without assuming the DMC's or SSP's specific architecture. With the EZ-diffusion model we test whether measurement models improve validity over raw behavioral indicators. Correlations between indicators and generating parameters serve as quantitative validity estimates; split-half reliability provides an independent assessment of indicator consistency.

For cross-task analyses, we generated data from two separate task administrations for each simulated participant, such that one or more parameters were correlated across tasks (e.g., controlled drift rate in Task 1 correlated with Task 2), testing whether parameter-level correlations transfer to indicator-level correlations. All simulations were conducted in R; code and analysis scripts are available in the ComputationalValidity R package (https://github.com/GidonFrischkorn/ComputationalValidity), and simulation data and results are archived at https://doi.org/10.17605/OSF.IO/6QNRV.

## Model-Based Validity Analysis

The simulations reveal four systematic patterns that challenge common measurement assumptions and demonstrate why explicit process models are necessary for validity assessment. For each phenomenon, we present results from both DMC and SSP to demonstrate that key findings generalize across architecturally distinct models.

### Selective Validity

![Comprehensive validity matrix showing correlations between all DMC parameters and behavioral indicators. The heatmap reveals selective validity: most indicators reflect a mixture of multiple processes (process impurity), but RT difference scores achieve relatively selective validity for control-related parameters.](../figures/manuscript/DMC_Equifinality){#fig-DMC-equifinality}

@fig-DMC-equifinality presents validity correlations between all five DMC parameters (controlled drift $\mu_c$, temporal dynamics $\tau$, automatic activation amplitude *A*, boundary separation *b*, and non-decision time $T_{er}$) and four indicator types (RT difference, accuracy difference, RT mean, accuracy mean). The dominant pattern is *process impurity*: Most behavioral indicators reflect a mixture of multiple cognitive processes. Accuracy difference scores substantially reflect both automatic activation amplitude (*A*: *r* = .57) and boundary separation (*b*: *r* = -.54), conflating control-related processes with response caution. RT mean scores primarily reflect non-decision time ($T_{er}$: *r* = .83) but also boundary separation (*b*: *r* = .43), conflating perceptual and motor processing with decision caution. Accuracy mean scores strongly reflect both boundary separation (*b*: *r* = .60) and controlled drift ($\mu_c$: *r* = .58), conflating caution with efficiency of processing relevant information. Interpreting any of these indicators as measuring a single construct oversimplifies the actual causal structure.

Against this backdrop of typical process impurity, RT difference scores are an exception: They show relatively *selective validity*. RT difference strongly reflects automatic activation amplitude (*A*: *r* = .78) and moderately reflects the temporal dynamics of automatic processing ($\tau$: *r* = .36), while showing minimal influences of other parameters (\|*r*\| $\leq$ .10 for $\mu_c$, *b*, and $T_{er}$). Both *A* and $\tau$ represent interference control mechanisms within the DMC framework: *A* reflects the strength of automatic response activation that must be controlled, while $\tau$ reflects the rate at which this activation is suppressed. RT difference thus validly measures attention control while remaining relatively uncontaminated by non-control processes.

These results show that different behavioral indicators are valid measures of different processes: If we want to measure interference control, RT difference scores provide valid measurement; if we want to measure response caution, accuracy mean scores are more appropriate (*b*: *r* = .60), with the caveat that these scores may still conflate multiple processes. No single indicator is universally "best". These differential validity patterns persist even with perfect reliability because they reflect the causal structure of how processes generate behavior.

![Comprehensive validity matrix showing correlations between all SSP parameters and behavioral indicators. As with DMC, RT difference scores achieve selective validity for the attention-control parameter (initial attentional width $sd_0$), while other indicators reflect mixtures of multiple processes.](../figures/manuscript/SSP_Equifinality){#fig-SSP-equifinality}

The SSP yields a convergent pattern (see @fig-SSP-equifinality). RT difference scores strongly reflect initial width of the attentional focus ($sd_0$: *r* = .70) – the SSP's attention-control parameter governing how broadly attention is initially spread across relevant and irrelevant features – with weaker influences from perceptual input (*P*: *r* = .27) and boundary separation (*b*: *r* = .21), and no influence of non-decision time ($T_{er}$: *r* = .00). As with DMC, process impurity characterizes most other indicators: Accuracy mean scores reflect perceptual input (*P*: *r* = .59), boundary separation (*b*: *r* = .55), and initial attentional width ($sd_0$: *r* = -.43). RT mean scores primarily reflect non-decision time ($T_{er}$: *r* = .76) and boundary separation (*b*: *r* = .50). Accuracy difference scores strongly reflect $sd_0$ (*r* = .72) but also boundary separation (*b*: *r* = -.39), again conflating attention-control processes with response caution.

The convergence across models highlights a key result: Despite different processing assumptions – the surge and decline of automatic processing in DMC versus attentional narrowing in SSP – both models identify RT difference scores as selectively valid for the attention-control parameter (automatic activation amplitude *A* and decline rate $\tau$ in DMC, initial attentional spotlight width $sd_0$ in SSP). This cross-model agreement suggests that the selective validity of RT difference scores for AC is a robust structural property of conflict tasks that does not depend on a particular model's assumptions.

### Reliability-Validity Dissociations

![Dissociation between reliability and validity of behavioral indicators for DMC. Panel A shows difference scores, Panel B shows mean scores; each column represents one generating parameter. Each data point represents one simulation. Higher variability is generated mostly through larger number of simulated trials. The dashed curve shows the theoretical upper bound on validity given reliability, $\sqrt{r_{xx}}$, derived from the classical attenuation formula when the generating parameter has perfect reliability. Data points above this curve indicate that validity exceeds the bound set by split-half reliability. Difference scores (A) show lower reliability but stronger validity for conflict-specific parameters, whereas mean scores (B) show near-perfect reliability but weak validity for these parameters.](../figures/manuscript/DMC_Reliability_Paradox){#fig-DMC-reliability}

@fig-DMC-reliability shows a dissociation between reliability and validity. Mean RT and accuracy scores show extremely high split-half reliability (*r* $\approx$ .99) but weak validity (\|*r*\| \< .30 for AC related parameters), whereas difference scores show moderate to low reliability (*r* $\approx$ .20–.70) but strong validity for control-related parameters (\|*r*\| \> .35 for *A* and $\tau$ for RT difference) that changes little with higher reliability. Notably, for some simulation conditions (i.e., parameter *A*, RT difference scores with low reliability) validity exceeds reliability – a pattern that challenges the conventional assumption that reliability sets an upper bound on validity.

The dissociation has a clear mechanistic explanation. Mean scores primarily reflect individual differences in parameters that affect behavior in congruent and incongruent conditions in the same way, whereas the parameters specific for AC processes have opposite effects in the two conditions: Strong control of the automatic process reduces interference in incongruent trials but reduces facilitation from the irrelevant stimulus dimension in congruent trials. Difference scores (subtracting congruent from incongruent performance) remove the contribution of processes common to both conditions, preserving variance related to AC-specific processes, but they do so at the cost of lower reliability. Optimizing reliability – a standard psychometric goal [@DraheimEtAl2021; @HedgeEtAl2018] – can thus reduce validity when it increases the influence of processes other than the target processes one aims to measure. The appropriate choice between reliable mean scores and noisier difference scores depends on what constructs the researcher aims to measure – an inherently theoretical question that psychometric properties alone cannot answer.

![Dissociation between reliability and validity for the SSP model. Panel A shows difference scores, Panel B shows mean scores; each column represents one generating parameter. The dashed curve shows the theoretical upper bound $\sqrt{r_{xx}}$. As with DMC, difference scores (A) show lower reliability but stronger validity for the attention-control parameter ($sd_0$), while mean scores (B) show near-perfect reliability but weaker validity for attention-control parameters.](../figures/manuscript/SSP_Reliability_Paradox){#fig-SSP-reliability}

The SSP simulations replicate the reliability-validity dissociation (@fig-SSP-reliability). Difference scores show moderate to low reliability but strong validity for initial attentional width ($sd_0$), the SSP's AC parameter. Mean scores again show near-perfect reliability but primarily reflect non-control parameters. The same mechanistic logic applies: In the SSP, the initial attentional spread ($sd_0$) determines how much irrelevant information influences processing. This effect operates in opposite directions across conditions – broader attention increases interference in incongruent trials but increases facilitation in congruent trials – so differencing removes condition-general variance while preserving control-specific variance. That both models produce the same dissociation for the same structural reason strengthens the conclusion that the reliability-validity trade-off is an inherent feature of conflict task measurement, not a model-specific artifact.

### Parameter Estimation with Cognitive Measurement Models: No Universal Solution

![Comparison of validity between behavioral indicators and EZ-Diffusion Model estimates for DMC parameters. Rows separate difference scores (top) from mean scores (bottom), for both behavioral indicators and EZ-DM estimates. Cognitive measurement models improve validity for some parameters but not others, showing no universal advantage.](../figures/manuscript/DMC_Intermediate_Modeling){#fig-DMC-modeling}

@fig-DMC-modeling compares validity for behavioral indicators versus EZ-Diffusion Model estimates [@WagenmakersEtAl2007]. Mean EZ-DM parameters (estimated from aggregate performance across conditions) are valid measures of corresponding parameters of the more complex process models. For example, EZ-DM boundary separation tracks the boundary parameter in the DMC well; EZ-DM non-decision time tracks DMC non-decision time, and EZ-DM drift rate reflects the DMC $\mu_c$ parameter. However, for conflict-specific parameters, behavioral indicators are more valid than EZ-DM parameters: RT difference scores show stronger validity for automatic activation amplitude (*A*) than any EZ-DM parameter, and also stronger validity than EZ-DM parameter differences between conditions, despite EZ-DM parameters controlling for speed-accuracy trade-offs that were often seen as a limitation for RT-based measures [@Draheim2019]. Parameters specific to DMC – particularly amplitude (*A*) and temporal dynamics ($\tau$) - show weak or negligible correlation with EZ-DM parameters because these processes are not represented in the simpler model.

![Comparison of validity between behavioral indicators and EZ-Diffusion Model estimates for SSP parameters. As with DMC, EZ-DM parameters validly reflect parameters that are represented in the EZ-DM (boundary, non-decision time) but fail to capture attention-control specific parameters ($sd_0$) that are not part of the EZ-DM.](../figures/manuscript/SSP_Intermediate_Modeling){#fig-SSP-modeling}

The SSP simulations confirm this pattern (@fig-SSP-modeling). EZ-DM boundary separation estimates validly reflect SSP boundary separation, and EZ-DM non-decision time tracks SSP non-decision time. However, for the attention-control parameter ($sd_0$), behavioral difference scores (RT difference: *r* $\approx$ .70; accuracy difference: *r* $\approx$ .72) substantially outperform all EZ-DM parameters, because $sd_0$ represents an attentional narrowing process that the EZ-DM does not model. Similarly, perceptual input (*P*) is only moderately captured by EZ-DM drift rate. The cross-model convergence reinforces the conclusion: Cognitive measurement models improve measurement for processes they represent but not for processes outside their scope.

We conclude that cognitive measurement models do not always improve validity relative to the use of behavioral indicators. Measurement models help when they adequately capture relevant processes but fail when the generative model includes processes the measurement model does not represent. The appropriate measurement model depends on the processes to be measured. Using an inadequate model may be worse than using well-chosen behavioral indicators, because the model imposes structure that misrepresents the processes of interest.

### Synthesis

These three analyses demonstrate systematic constraints on behavioral measurement that emerge consistently from both DMC and SSP. Validity is parameter-specific: Different behavioral indicators measure different processes, even from the same task. Claiming that a task measures "attention control" requires specifying which indicator is obtained from the task and demonstrating that this indicator captures the intended mechanisms [@FrischkornOberauer2025]. Increasing reliability may reduce validity when it is achieved by changes to the task or the indicator that increase the influence of nuisance processes. Cognitive measurement models are suited for measuring those processes that are included in the models but are ill suited for measuring other processes. These constraints cannot be resolved through psychometric refinement alone – they require explicit process theory. Critically, both AC models converge on the same key conclusion: RT difference scores are selectively valid indicators of AC processes, regardless of whether one assumes the DMC's temporal conflict dynamics or the SSP's attentional narrowing mechanism.

## Parameter Overlap and Cross-Task Validity

Having established which indicators validly reflect which parameters from a formal model, we now turn to the question how validity influences correlations of indicators across tasks. Even when cognitive processes are perfectly shared across tasks, behavioral indicator correlations will likely underestimate the parameter-level correlations – a phenomenon that explains why cross-task convergent validity is typically moderate.[^5] This underestimation will happen to the extent that the behavioral indicator is influenced not only by the processes that are shared between tasks but also by other processes that are not shared. When an indicator is highly valid for a single parameter – as RT difference scores are for attention-control parameters – correlations between parameters should transfer well to correlations between the indicators. In contrast, when an indicator is process-impure, reflecting multiple parameters of which only some are shared across tasks or correlate, correlations between the indicators are attenuated relative to correlations between the parameters. This principle follows from the differential validity documented above (@fig-DMC-equifinality and @fig-SSP-equifinality).

[^5]: When an indicator reflects multiple parameters but only one is shared across tasks, variance contributed by uncorrelated parameters dilutes the observed cross-task correlation relative to the true parameter-level correlation.

![Correlation transfer from model parameters to behavioral indicators for DMC simulations. Each column represents a single parameter correlated across tasks (all other parameters uncorrelated). Rows separate difference scores (top) from mean scores (bottom). Response Time indicators are shown in orange, Accuracy in blue. Lines show linear regression fits with 95% confidence bands; individual simulation results are shown as faint points. The dashed identity line marks perfect transfer. Points consistently fall below this line, indicating that even strong parameter-level correlations transfer only partially to indicator-level correlations.](../figures/manuscript/DMC_Correlation_Transfer){#fig-DMC-correlation}

@fig-DMC-correlation presents results from simulations where we generated data from two independent "tasks" (separate applications of the same DMC model), inducing known correlations between corresponding parameters across tasks. For example, we might set controlled drift in Task 1 to correlate *r* = .70 with controlled drift in Task 2, simulating individuals who show consistent controlled processing efficiency across tasks. In each simulation only one parameter was positively correlated across tasks; all other parameters were sampled independently for the two tasks. We then computed cross-task correlations among behavioral indicators and compared them to the generating parameter correlations. @fig-DMC-correlation shows results for RT and accuracy means as well as difference scores.

The results confirm that the degree of correlation transfer depends on the validity of the indicator for the correlated parameter. Looking across the individual parameter panels, RT difference scores show positive correlation transfer for both *A* and $\tau$ – the two AC-relevant parameters for which RT differences have high validity. When these parameters correlate at *r* = .60 to .80 across tasks, cross-task correlations of RT difference scores approximate these parameter correlations with only modest attenuation. Because RT difference scores are relatively process-pure for AC parameters, cross-task correlations of RT difference scores provide a meaningful window into shared cognitive processes.

By contrast, when only a single parameter is correlated across tasks and the indicator reflects multiple parameters of which some might be uncorrelated, attenuation is more pronounced. For example, cross-task correlations of mean accuracy do not approximate the parameter-level correlations of $\mu_c$ well, although mean accuracy does correlate with $\mu_c$ substantially (*r* = .60). But because mean accuracy also reflects variations in *boundary separation* which are uncorrelated across tasks, the transfer of $\mu_c$ correlations to indicator correlations is strongly attenuated. The degree of attenuation of cross-task correlation between parameters is thus proportional to how much of the indicator's variance is explained by the correlated parameter versus other parameters.

Note that this design represents a best-case scenario: Both tasks share the identical processing architecture, differing only in parameter values. When real tasks engage overlapping but non-identical process architectures – as is likely for flanker versus Simon or Stroop tasks – correlation transfer would be further attenuated.

![Correlation transfer from model parameters to behavioral indicators for SSP simulations. Each column represents a single parameter correlated across tasks (all other parameters uncorrelated). Rows separate difference scores (top) from mean scores (bottom). Response Time indicators are shown in orange, Accuracy in blue. Lines show linear regression fits with 95% confidence bands; individual simulation results are shown as faint points. The same attenuation pattern as for DMC emerges, demonstrating that correlation transfer failure generalizes across model architectures.](../figures/manuscript/SSP_Correlation_Transfer){#fig-SSP-correlation}

The SSP simulations replicate and extend these patterns (see @fig-SSP-correlation). Correlation transfer again depends on validity: Both accuracy and RT difference scores approximate the cross-task correlations of $sd_0$ – the parameter for which they are highly valid (*r* = .70 for RT, *r* = .72 for accuracy) – but fail to approximate the correlations of perceptual input *P*, for which their validity is low. Conversely, mean scores show some transfer for *P* correlations, consistent with their moderate validity for *P* (accuracy mean: *r* = .59), but do not approximate $sd_0$ correlations at all.

Process purity governs the degree of attenuation even where validity is established. Mean accuracy is moderately valid for *P* but also reflects boundary separation (*r* = .55) and $sd_0$ (*r* = -.43). When only *P* is correlated across tasks, the variance contributed by uncorrelated *b* and $sd_0$ attenuates the observed correlation – the same mechanism that attenuates mean accuracy correlations for $\mu_c$ in DMC. The cross-model convergence confirms that the principle holds generally: Correlation transfer depends jointly on validity and process purity, regardless of the specific model architecture.

These results reveal two complementary consequences of process impurity for interpreting cross-task correlations. On the one hand, weak cross-task correlations do not necessarily imply that the construct we aim to measure does not generalize across tasks – they may reflect measurement limitations even when the underlying processes are perfectly shared. On the other hand, high cross-task correlations of process-impure indicators might also reflect shared non-target processes rather than the processes of interest. For instance, if mean proportion correct were used to measure AC, a substantial cross-task correlation is observed if the controlled drift rate (DMC) or perceptual processing rate (SSP) is correlated between tasks, while attention-control parameters are not.

These patterns are not contradictory but follow from the same principle: Behavioral correlations are interpretable as evidence about parameter-level relationships only when there is clear evidence for the validity of the indicator for the parameter in question. When this condition is met – as it is for RT difference scores and attention-control parameters – cross-task correlations become meaningful tests of whether the construct generalizes across tasks.

These patterns have concrete implications for the current debate around the measurement of attention control (AC) processes. First, RT difference scores – despite their low reliability – are not only the most selectively valid indicators for AC-related parameters across both DMC and SSP, but also show adequate correlation transfer when AC processes are shared across tasks. Given this, the weak inter-task correlations of RT difference scores documented in the literature [@ReyMermetEtAl2018; @EnkaviEtAl2019] can be interpreted as evidence that AC, at least in terms of the model parameters evaluated here, is not a unitary construct. Second, the strategy of improving reliability by using mean scores or aggregated measures that combine speed and accuracy [e.g., accuracy under response deadlines; @DraheimEtAl2021] trades AC validity for reliability: Mean RT and accuracy primarily reflect cognitive processing parameters that are not unique to conflict tasks and do not characterize AC processes involved in conflict resolution, namely general rate of evidence accumulation, non-decision time, and boundary separation. Stronger inter-task correlations achieved this way thus likely reflect shared attributes of general decision processes rather than shared efficiency of attention control [@LoefflerEtAl2024; @FrischkornOberauer2025]. Third, applying an intermediate measurement model such as the EZ-Diffusion Model as proposed by @Yanguez2024 also does not resolve the AC measurement problem: EZ-DM parameters validly track the non-AC parameters they represent (boundary separation, non-decision time, overall drift rate) but show weaker validity for the conflict-specific parameters than RT difference scores do. Differences between diffusion-model parameters for congruent vs. incongruent trials, as tried by @ReyMermet2025, still do not achieve the same validity as RT differences. None of these strategies – optimizing reliability, or applying an intermediate measurement model – substitutes for identifying indicators whose validity for AC processes has been established through simulation-based analysis.

In sum, our proposed framework builds on the premise that individual differences in cognitive processes should be located at the level of model parameters [@Oberauer2016; @FrischkornEtAl2022], not task scores. Rather than defining "attention control" as the common factor among conflict task scores, we should define it through individual differences in process parameters that characterize the processes that control attention. This approach grounds construct definition in explicit process theory rather than empirical covariance.

## Measurement Implications

Strong construct claims – asserting that a task measures a specific attribute of a cognitive mechanism or process – require explicit process theory. Without it, correlational patterns cannot establish what processes drive performance. Traditional psychometric practice reasons that if *N* tasks form a coherent factor, the factor reflects what the tasks have in common, and researchers infer the shared process from an informal task analysis. Our framework shows why this informal analysis is insufficient: An indicator may indeed capture the process of interest, but it also captures other processes, and the factor may primarily reflect the latter. For example, @FrischkornOberauer2025 decomposed performance in the antisaccade task, a popular measure of cognitive inhibition, into several component processes and demonstrated that inhibition of reflexive saccades – the process commonly assumed to be measured by the antisaccade task – accounts for only a small fraction of performance variance, whereas other parameters, such as the procedural binding strength and perceptual speed, account for much more. These other parameters are likely shared across many tasks and could drive factor coherence without the factor actually reflecting inhibition as intended.

These findings point to two complementary approaches for grounding individual-differences claims in explicit process theory. The first – simulation-based validity assessment, which is the focus of this paper – determines *which* behavioral indicators can measure *which* processes. It uses existing generative models to simulate data, computes the same behavioral indicators that empirical researchers use, and assesses how strongly those indicators correlate with the generating parameters. This requires no new data collection: once established, validity assessments based on a generative model transfer to any study using the same indicators in the same paradigm.

The second approach – model fitting – *estimates parameters directly* by fitting process models to empirical data for each individual [@Vandekerckhove2014; @FrischkornEtAl2022]. Both approaches deliver a measurement whose correlation with the true parameter can be assessed through simulation: for simulation-based validity assessment, this correlation is the validity estimate we report throughout this paper; for model fitting, the analogous quantity is the parameter recovery correlation obtained from recovery simulations. Neither approach is uniformly superior. When the parameter recovery correlation exceeds the validity correlation of the best behavioral indicator, researchers should estimate parameters directly; when the reverse holds, they should use the most valid behavioral indicator. The choice is thus an empirical question, answerable by comparing the two correlations for each parameter and paradigm. Most existing studies can apply simulation-based validity assessment immediately; model fitting becomes attractive when adequate data and suitable model-fitting routines are available.

## Model-Relativity and Generalizability

One natural objection to the framework we propose is that validity claims become model-dependent: Different models may yield different conclusions about which indicators are valid. From our perspective, this model-relativity is a strength, not a limitation. All validity claims depend on theoretical commitments [@Borsboom2004]. In traditional psychometrics these are the often tacit assumptions underlying the informal task analysis through which task scores and factors are interpreted. Interpreting the flanker effects as measuring “conflict resolution” rests on the assumption that incongruent stimuli create response conflict and that some attention-control processes are engaged in resolving that conflict. Interpreting complex span scores as “working memory capacity” rests on the assumption that a combination of memory and processing demands specifically taxes capacity limits. These interpretations rest on implicit theories about cognitive processes. Making theoretical assumptions explicit does not create model-dependence; it makes pre-existing commitments visible and testable.

We do not have the choice between validity claims that depend on theoretical assumptions and validity claims that don't; we only have the choice between validity claims that depend on theoretical assumptions that are more or less precisely and explicitly specified. The more they are, the better we can evaluate, revise, and improve them. This observation reflects a fundamental insight from the philosophy of science: All scientific inference is conducted through models, and model-dependence is an epistemic condition of inquiry, not a special liability of formal approaches [@Giere1988; @Suppes1962]. Making theoretical commitments explicit converts untestable assumptions into testable ones [@Navarro2021; @GuestMartin2021], which is what distinguishes scientific model-dependence from arbitrary convention [@vanRooijBaggio2021].

All validity claims are thus conditional: "*Given* theoretical assumptions X, indicator Y validly measures parameter Z with validity W." Conditionality should not be confused with circularity. Circular reasoning uses its conclusion as a premise – the correlational pattern that is supposed to validate a measure is simultaneously used to define the construct. Conditional reasoning, by contrast, makes explicit what must be true for the conclusion to hold, and those assumptions can be tested and rejected independently. This conditionality is epistemic honesty, not a weakness. The simulations reported above demonstrate that some validity patterns emerge from both DMC and SSP despite their different processing assumptions – in particular, RT difference scores validly reflect attention-control parameters under both models. This illustrates a case where validity is not tied to the assumptions of one particular model but rather to a set of more general assumptions that these models share; in the present case, the assumptions that decisions rely on evidence accumulation towards a boundary, and that the irrelevant feature dimension contributes strongly to the accumulated evidence early in the trial and is reduced over time. In other cases, the validity of a behavioral indicator will differ between competing models. In those cases it is possible to adjudicate between the models through competitive model fitting and experimental manipulations for which the models make different qualitative predictions (e.g., about the shape of RT distributions or the direction of congruency effects).

Our approach shares its philosophical foundation with @KvamAlaukik2025, who propose replacing operational definitions across psychology with "relational definitions" grounded in computational models, and also with the cognitive psychometrics tradition [@Vandekerckhove2011; @Vandekerckhove2014; @RouderLu2005; @FrischkornEtAl2022], which locates individual differences at the parameter level by fitting generative models to empirical data. Whereas @KvamAlaukik2025 and the cognitive psychometrics approach emphasize joint hierarchical parameter estimation across tasks – the model-fitting route pioneered by @Vandekerckhove2014 - we propose a complementary route that does not require model fitting. Simulation-based validity assessment allows researchers to identify which behavioral indicators are valid for which model parameters, and how strongly, using only existing generative models. This route is often easier to use because it works with existing behavioral data and requires no new data collection or model-fitting expertise. As outlined above, the two routes are complementary: Simulation-based validity assessment establishes which indicators to use and what their measurement limitations are, while model fitting provides individual-level parameter estimates when adequate data and suitable model-fitting routines are available.

## Constructs Follow Theory

The framework developed here has a deeper implication that extends beyond the question of which indicators validly measure which processes: If we want to measure an existing construct such as intelligence or working-memory capacity, we might find that the best available theory of behavior in established tests of that construct makes no reference to a single mechanism or process that represents the construct. When that is the case, then within that theory there is no attribute that causally underlies variation in observed behavior for that construct. Following @Borsboom2004, valid measurement of an attribute requires that the attribute exists and causally produces variation in the measurement outcome. Hence, in such a case the construct cannot be validly measured within that theoretical framework. This does not mean the construct is useless, but it means that measuring it does not measure an attribute of the mind.

Consider, for example, the mutualism theory of intelligence [@vanderMaasEtAl2006]. In this theory, the positive manifold of cognitive abilities – the finding that diverse cognitive tests tend to correlate positively – arises not from a single general ability factor (*g*) that causally influences all test scores, but from mutually reinforcing interactions among basic cognitive processes. Each process benefits from the efficiency of others, creating emergent positive correlations across tasks. Within this framework, we can measure the individual processes and their interaction strengths, but "intelligence" is not an entity that causes differences in observed behavior – it emerges from differences in the processes and their interconnectedness. As @vanderMaasEtAl2014 argue, under mutualism *g* should be understood as a formative index – analogous to a stock market index – that is useful as a summary description of a person's cognitive ability, and predictive for many outcomes, but it does not measure a latent common cause. Just as companies in a stock index correlate positively through their interactions in a complex economic ecosystem rather than through dependence on a single latent factor, cognitive abilities in the mutualism model correlate through reciprocal facilitation. IQ scores remain scientifically respectable as index variables, but their validity claim shifts from measurement of a latent attribute to computation of a summary index [@vanderMaasEtAl2014; @SijtsmaEtAl2024].

A similar situation arises for working memory capacity (WMC). Complex span tasks are routinely used to measure WMC as a unitary construct [@Engle2002; @EngleEtAl1999], yet the best available process models of complex span performance – such as SOB-CS [@OberauerLewandowsky2011] – have no single parameter corresponding to WMC. Instead, multiple parameters that quantify attributes of different mechanisms jointly determine performance: Encoding strength, interference from distractor processing, distinctiveness of memory representations, and the effectiveness of removal processes. If we take the SOB-CS model seriously as one of our best theoretical accounts of complex span performance, WMC as traditionally conceived is not a measurable unitary construct. Individual differences in complex span performance reflect a mixture of differences in several distinct processes, none of which individually corresponds to "working memory capacity." This would mean that working memory capacity is not an explanatory construct but rather a descriptive term that refers to the performance limits of people working on working-memory tasks.

Similar to validity being model dependent, this is not a limitation of the framework we propose but a desirable feature. It forces researchers to confront the theoretical status of their constructs. When the best available theory dissolves a construct into component processes, we have the choice between three options. One is to conceptualize the construct as a formative index, a summary score that is useful for predicting practically relevant outcomes such as academic success or recovery from drug addiction. The second option is to develop theories in which the construct of interest appears as a causal parameter. The third option is to abandon the construct and instead measure individual differences in terms of the component processes that, according to our best theory, cause the behavior that we used to summarize through the construct.

This principle extends beyond the specific examples of intelligence and working memory. @vanBorkEtAl2024 demonstrate that reconceptualizing psychological constructs as networks – systems of causally interacting components rather than effects of a latent common cause – fundamentally changes what measurement targets are available. For network constructs such as depression, they identify five distinct targets: the network structure itself, the current state (number of active nodes), the expected state given the network's dynamics, the vulnerability of the system's dynamical landscape, and the stable state the system occupies. Each target requires different measurement approaches and different psychometric criteria. Normative criteria derived from latent variable theory – such as local independence, high factor loadings, or measurement invariance – do not automatically transfer to network constructs, because these criteria presuppose a specific conceptualization of the measurement target that the network framework does not share. This illustrates the broader point: construct theory determines not only which indicators are valid, but what the measurement target itself is. Different theories of the same domain yield different quantities to be measured, each with their own validity requirements.

## Conclusion

*Validity is fundamentally a theoretical achievement, not a psychometric property.* It requires explicit commitments about what cognitive processes operate and how they generate observable behavior [@Borsboom2004; @KvamAlaukik2025]. Decades of psychometric refinement have not resolved what cognitive tasks measure because the problem was never primarily psychometric – it is theoretical.

The framework developed here addresses this by making theoretical commitments explicit and testable. Using generative computational models, we can assess the validity of behavioral indicators independently of task correlations and generate falsifiable predictions about indicator-process relationships. Applied to the measurement of attention control, this framework yielded a key positive finding: RT difference scores are selectively valid indicators of AC parameters under both the DMC and SSP, providing a measurement basis that is robust across different processing assumptions. At the same time, the simulations revealed why many commonly used indicators fail to measure what they are intended to measure: Process impurity means that most behavioral indicators conflate multiple cognitive processes, leading to reliability-validity dissociations and attenuated correlation transfer for process-impure indicators. These patterns illustrate why validity assessment requires explicit process theory, not just psychometric refinement.

These contributions come with important constraints. Our simulations focused on conflict tasks; although the structural phenomena we documented arise from general features of multi-process measurement and should emerge wherever multiple parameters jointly determine behavioral indicators, demonstrating this for other domains – working memory, decision-making, learning – remains an important next step. The framework requires computational modeling expertise that is not yet widespread in individual-differences research, and well-tested generative models exist for only a subset of cognitive domains. Validity assessment through simulation establishes measurement potential under idealized conditions; whether these patterns hold in empirical data depends on model adequacy, which must be evaluated for the class of tasks and paradigms under study. Moreover, the models we used assume stationary processes and independent trials – real cognitive performance involves learning, fatigue, and strategic shifts that may alter process-indicator relationships. These limitations define the framework's current boundaries, not its ultimate reach: As generative models are developed for new domains and tested against empirical data, the approach becomes progressively more applicable.

@CronbachMeehl1955 recognized seven decades ago that establishing construct validity rests on theory testing. Computational cognitive models now make it possible to translate this theoretical insight into practice, providing the explicit process theories that their framework required but their era could not supply. Advancing validity in measuring cognition requires not better correlations, but better theories of how cognition generates behavior.

{{< pagebreak >}}

::: {#refs}
:::

{{< pagebreak >}}

# Appendix: Supplementary Methods and Results {.appendix}

## Figure A3: Model Validation – Empirical Pattern Alignment for DMC and SSP

![**Panel A:** Distribution of RT congruency effects (incongruent minus congruent mean RT, in ms) across simulated participants under the DMC (left) and SSP (right), based on the recovery simulations (*N* = 100, 100 trials per condition, 250 replications). The grey shaded band indicates the typical empirical range for flanker and Simon tasks (20–100 ms). Both models produce predominantly positive congruency effects within this range, confirming that the simulation parameter space generates behavioral data consistent with empirical conflict-task observations. **Panel B:** Distributions of mean RT for congruent (blue) and incongruent (red) conditions. Both models produce the expected pattern of slower responses in incongruent than congruent trials, with realistic RT ranges. **Panel C:** Distributions of mean accuracy for congruent (blue) and incongruent (red) conditions. Both models produce higher accuracy in congruent than incongruent trials, consistent with empirical conflict-task data [@UlrichEtAl2015; @WhiteEtAl2011; @WhiteEtAl2018].](../figures/manuscript/DMC_SSP_ModelValidation.png){#fig-modelvalidation}
