---
title: "Establishing validity from a computational modeling perspective"
subtitle: "An example focussing on inhibition."
author: "Gidon T. Frischkorn, Ricardo Rebmann, & Klaus Oberauer"
format: html
editor: visual
bibliography: references.json
embed-resources: TRUE
editor_options: 
  chunk_output_type: console
---

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = FALSE,
  output = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  collapse = TRUE
)
```

```{r}
# use relative paths to load & save data
pacman::p_load(here, SimDesign, tidytable, data.table, ggplot2, ComputationalValidity)
# source(here("scripts","LoadResultsFiles.R"))
```

## Establishing validity from a computational modeling perspective: An example focusing on inhibition

Following commonly used definitions of validity and propositions of validity researchers, establishing the validity of a newly developed measure or test requires its integration in the nomological network of psychological constructs [@cronbachConstructValidityPsychological1955] and providing sufficient evidence for its interpretation and practical use [@messickValidity1987]

## Conflating theoretical with psychometrics questions

## Validity from a computational modeling perspective

### Computational Models implement how processes cause changes in observed behavior

### Identifying suitable compuational models

## Identifying valid behavioral performance measures for inhibition processes

Here we focus on computational models for traditional inhibition tasks such as the Stroop [@stroopStudiesInterferenceSerial1935], Flanker [@eriksenEffectsNoiseLetters1974], or Simon [@simonEffectCompatibilitySR1981] task. These are particularly relevant as the traditional indicator for inhibition from these tasks - that is the difference of reaction times from correct responses between incongruent and congruent conditions- has been criticized for its bad psychometric properties [@draheimReactionTimeDifferential2019; @vonbastianAdvancingUnderstandingIndividual2020]. The main point of criticism was that reaction times are affected by speed-accuracy trade offs and tend to be unreliable, in particular when considering experimental effect measures.

### Simulation I: Diffusion Model for Conflict Tasks

#### Behavioral Effects

```{r avg-behavioral-effects}
dmc_effects_behavior <- dmc_recovery_behavior %>%
  summarize(mean = mean(value),
            .by = c(nRep, SampleSize, nTrials, measure, indicator)) %>%
  mutate(mean = case_when(measure == "RT" ~ mean * 1000,
                          TRUE ~ mean * 100)) %>% 
  summarize(mean = mean(mean),
            min = min(mean),
            max = max(mean),
            .by = c(SampleSize, nTrials, measure, indicator)) %>% 
  pivot_wider(names_from = measure,
              values_from = c(mean, min, max)) %>% 
  filter(indicator %in% c("mean","difference")) %>% 
  arrange(indicator, SampleSize, nTrials) %>% 
  select(indicator, SampleSize, nTrials,
         mean_RT, min_RT, max_RT,
         mean_PC, min_PC, max_PC)
```

```{r}
#| label: tbl-FX-DMC
#| tbl-cap: "Summary of the average performance and conflict effects in the data simulated from the DMC"
knitr::kable(dmc_effects_behavior, digits = 2)
```

```{r}
dmc_effects_ezDM <- dmc_recovery_ezDM %>%
  summarize(mean = mean(value),
            .by = c(nRep, SampleSize, nTrials, measure, indicator)) %>%
  mutate(mean = case_when(measure == "non_dec" ~ mean * 1000,
                          TRUE ~ mean)) %>% 
  summarize(mean = mean(mean),
            min = min(mean),
            max = max(mean),
            .by = c(SampleSize, nTrials, measure, indicator)) %>% 
  pivot_wider(names_from = measure,
              values_from = c(mean, min, max)) %>% 
  filter(indicator %in% c("mean","difference")) %>% 
  arrange(indicator, SampleSize, nTrials) %>% 
  select(indicator, SampleSize, nTrials,
         mean_drift, min_drift, max_drift,
         mean_boundary, min_boundary, max_boundary,
         mean_non_dec, min_non_dec, max_non_dec)
```

```{r}
#| label: tbl-FX-ezDM-DMC
#| tbl-cap: "Summary of the observed condition differerences and conflict effects in the data simulated from the DMC"
knitr::kable(dmc_effects_ezDM, digits = 2)
```

```{r merge-effects}
dmc_effects_behavior <- dmc_recovery_behavior %>%
  summarize(mean = mean(value), sd = sd(value),
            .by = c(nRep, SampleSize, nTrials, measure, indicator)) %>%
  mutate(mean = case_when(measure == "RT" ~ mean * 1000,
                          TRUE ~ mean * 100))
dmc_effects_ezDM <- dmc_recovery_ezDM %>%
  summarize(mean = mean(value), sd = sd(value),
            .by = c(nRep, SampleSize, nTrials, measure, indicator)) %>%
  mutate(mean = case_when(measure == "non_dec" ~ mean * 1000,
                          TRUE ~ mean))

avg_effects_dmc <- rbind(dmc_effects_behavior,dmc_effects_ezDM)

dmc_reliability <- dmc_recovery_reliability %>%
  left_join(avg_effects_dmc)

dmc_parRecovery <- dmc_recovery_parRecovery %>%
  left_join(avg_effects_dmc)
```

#### Reliability

```{r}
ggplot(data = dmc_reliability,
       aes(x = as.factor(nTrials), y = reliability, fill = as.factor(SampleSize))) +
  facet_grid(indicator ~ measure) +
  ylim(-1,1) +
  geom_hline(yintercept = 0, linewidth = 1) +
  geom_hline(yintercept = 0.7, linewidth = 1, linetype = "dotted", color = "green") +
  geom_hline(yintercept = 0.9, linewidth = 1, linetype = "dotted", color = "darkgreen") +
  geom_boxplot() +
  labs(x = "Number of Trials in each Condition", y = "Estimated Reliability",
       fill = "Sample Size",
       title = "Reliability estimates (Split-Half correlation) of performance indicators") +
  theme_gf()

ggplot(data = dmc_reliability %>% filter(indicator == "difference", SampleSize == "N = 100"),
       aes(x = sd, y = reliability, color = as.factor(nTrials))) +
  facet_grid(nTrials ~ measure, scales = "free") +
  ylim(-1,1) +
  geom_point(alpha = 0.1, color = "darkgrey") +
  geom_smooth(method = "lm") +
  labs(x = "Standard Deviation of Conflict Effect", y = "Estimated Reliability",
       color = "Number of Trials") +
  theme_gf()
```

#### Recovery of DMC parameters

```{r}
ggplot(data = dmc_parRecovery %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("RT","PC")),
       aes(x = genPar, y = rec, fill = as.factor(nTrials))) +
  facet_grid(measure ~ indicator) +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_boxplot() +
  labs(x = "Generating DMC Parameter", y = "Recovery",
       fill = "Sample Size") +
  coord_cartesian(ylim = c(-1,1)) +
  theme_gf()

ggplot(data = dmc_parRecovery %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("RT","PC"),
                indicator %in% c("difference")),
       aes(x = mean, y = rec, color = as.factor(SampleSize))) +
  facet_grid(measure ~ genPar) +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  labs(x = "Mean Conflict Effect in the Sample", y = "Recovery",
       fill = "Sample Size") +
  coord_cartesian(ylim = c(-1,1)) +
  theme_gf()


ggplot(data = dmc_parRecovery %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("RT","PC"),
                indicator %in% c("difference")),
       aes(x = sd, y = rec, color = as.factor(SampleSize))) +
  facet_grid(measure ~ genPar, scale = "free") +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm") +
  labs(x = "SD of Conflict Effect in the Sample", y = "Recovery",
       fill = "Sample Size") +
  coord_cartesian(ylim = c(-1,1)) +
  theme_gf()


ggplot(data = dmc_parRecovery %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("drift","boundary","non_dec")),
       aes(x = genPar, y = rec, fill = SampleSize)) +
  facet_grid(measure ~ indicator) +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_boxplot() +
  coord_cartesian(ylim = c(-1,1)) +
  labs(x = "Generating DMC Parameter", y = "Recovery",
       fill = "Sample Size") +
  theme_gf()


ggplot(data = dmc_parRecovery %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("RT","PC"),
                SampleSize == "N = 100"),
       aes(x = genPar, y = rec_corrected, fill = as.factor(nTrials))) +
  facet_grid(measure ~ indicator) +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_boxplot() +
  ylim(-1,1) +
  coord_cartesian(ylim = c(-1,1)) +
  labs(x = "Generating DMC Parameter", y = "Recovery (Corrected for Reliability)",
       fill = "Sample Size") +
  theme_gf()


ggplot(data = dmc_parRecovery %>%
         filter(genPar %in% c("A","b","muc","non_dec","tau"),
                measure %in% c("drift","boundary","non_dec"),
                SampleSize == "N = 100"),
       aes(x = genPar, y = rec_corrected, fill = as.factor(nTrials))) +
  facet_grid(measure ~ indicator) +
  geom_hline(yintercept = 0, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 0.3, color = "darkred", linetype = "dotted") +
  geom_hline(yintercept = -0.3, color = "darkred", linetype = "dotted") +
  geom_boxplot() +
  ylim(-1,1) +
  coord_cartesian(ylim = c(-1,1)) +
  labs(x = "Generating DMC Parameter", y = "Recovery (Corrected for Reliability)",
       fill = "Sample Size") +
  theme_gf()
```

### Simulation II: Shrinking Spotlight Model of Attention

## Appendix
